---
title: "Final Project Graphs"
subtitle: "Data Visualization"
author: "Ryan Shanker"
output:
  html_document:
    toc: true
    toc_float: true
    highlight: "tango"
---

These graphs were created using the three following datasets:

1. A dataset of RAPTOR statistics for modern NBA players for the 2014-2021 seasons from FiveThirtyEight's Github (https://github.com/fivethirtyeight/data/blob/master/nba-raptor/modern_RAPTOR_by_player.csv)

2. A table of all NBA player salaries from the 2021-22 season to the future 2024-25 season from HoopsHype (https://hoopshype.com/salaries/players/)

3. Datasets of the win-loss records for the teams in each conference for the first 25 games of the 2021-22 season
from Basketball Reference (https://www.basketball-reference.com/leagues/NBA_2022_standings.html)

## Graph 1
```{r echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(tidyverse)
library(janitor)
library(dplyr)
library(stringi)

nbaraptor <- read_csv("modern_RAPTOR_by_player.csv")
nbasalaries <- read_csv("nbasalaries2021.csv")
winpct <- read_csv("winpct.csv")

nbasalaries <- nbasalaries %>% 
  janitor::clean_names()

nbasalaries <- subset (nbasalaries, select = -x1)
winpct <- subset(winpct, select = -...1)

names(nbaraptor)[1] <- 'player'
names(nbasalaries)[3] <- 'salary2021_22'

nbasalaries$player <- stri_trans_general(nbasalaries$player, "Latin-ASCII")

nbaraptor2021 <- nbaraptor %>% 
  filter(season == 2021 & mp >= 500)

nbasalaries$salary2021_22 = as.numeric(gsub("[\\$,]", "", nbasalaries$salary2021_22))

nbaraptordata <- subset(nbaraptor2021, player %in% nbasalaries$player)
nbasalariesdata <- subset(nbasalaries, player %in% nbaraptordata$player)

fullnbadata <- merge(nbaraptordata, nbasalariesdata, by = 'player')
```

```{r echo = FALSE, message = FALSE}
raptorscaler <- function(player2) {
  
  filtered <- fullnbadata %>% 
  filter(player == player2)
  
  x = ((filtered$raptor_total - mean(fullnbadata$raptor_total))/(sd(fullnbadata$raptor_total)))
  
  return(x)
}

warscaler <- function(player2) {
  
  filtered <- fullnbadata %>% 
  filter(player == player2)
  
  x = ((filtered$war_total - mean(fullnbadata$war_total))/(sd(fullnbadata$war_total)))
  
  return(x)
}

boxscaler <- function(player2) {
  
  filtered <- fullnbadata %>% 
  filter(player == player2)
  
  x = ((filtered$raptor_box_total - mean(fullnbadata$raptor_box_total))/(sd(fullnbadata$raptor_box_total)))
  
  return(x)
}

salaryscaler <- function(player2) {
  
  filtered <- fullnbadata %>% 
  filter(player == player2)
  
   y = ((filtered$salary2021_22 - mean(fullnbadata$salary2021_22))/(sd(fullnbadata$salary2021_22)))
  
  return(y)
}

```

```{r echo = FALSE, message = FALSE}
fullnbadata <- fullnbadata %>% 
  mutate(RAPTOR_scaled = raptorscaler(player))

fullnbadata <- fullnbadata %>% 
  mutate(WAR_scaled = warscaler(player))

fullnbadata <- fullnbadata %>% 
  mutate(RAPTOR_box_scaled = boxscaler(player))

fullnbadata <- fullnbadata %>% 
  mutate(salary_scaled = salaryscaler(player))

fullnbadata <- fullnbadata %>% 
  mutate(difference = ((RAPTOR_box_scaled + WAR_scaled + RAPTOR_scaled))/3 - salary_scaled)

fullnbadata <- fullnbadata %>% 
  mutate(gooddeal = ifelse(difference >= 1, "Well Underpaid",
                           ifelse(difference <= -1, "Well Overpaid",
                                  ifelse(difference < 1 & difference >= 0.50, "Slightly Underpaid",
                                         ifelse(difference > -1 & difference <= -0.50, "Slightly Overpaid",
                                                "Appropriately Paid")))))
```

```{r echo = FALSE, message = FALSE}
deallabels <- c("Well Overpaid", "Slightly Overpaid", "Appropriately Paid", "Slightly Underpaid",
                   "Well Underpaid")

dealfrac <- c(rep(0, 5))

for (ii in 1:5) {
  filterer <- fullnbadata %>% 
    filter(gooddeal == deallabels[ii])
  
  dealfrac[ii] = nrow(filterer)/nrow(fullnbadata)
}

pielabels <- paste0(deallabels, " (", round(100 * dealfrac, 2), "%) ")

dealdata <- data.frame(deallabels, dealfrac)
```

```{r echo = FALSE, message = FALSE}
pie(dealdata$dealfrac, labels = pielabels, col = c('red',
                                                   'orange',
                                                   'yellow',
                                                   'light green',
                                                   'green'),
    main = "Pie Chart of NBA Player Salaries, Judged by Our Scaling Method")
```

## Graph 2
```{r echo = FALSE, message = FALSE}
data <- lapply(fullnbadata, unique)
teamlist <- data$team

badcontractjudge <- c(rep(0, 30))
goodcontractjudge <- c(rep(0, 30))

for (jj in 1:30) {
  filterer <- fullnbadata %>% 
    filter(team == teamlist[jj] & (gooddeal == "Well Overpaid" | gooddeal == "Slightly Overpaid"))
  
  badcontractjudge[jj] = nrow(filterer)
}

for (jj in 1:30) {
  filterer <- fullnbadata %>% 
    filter(team == teamlist[jj] & (gooddeal == "Well Underpaid" | gooddeal == "Slightly Underpaid"))
  
  goodcontractjudge[jj] = nrow(filterer)
}

contractjudge <- goodcontractjudge - badcontractjudge

contractdata <- data.frame(teamlist, contractjudge)
names(contractdata)[1] <- 'team'

teamcolors <- c("#E03A3E", "#007A33", "#777D84", "#CE1141", "#00788C", 
                "#860038", "#00538C", '#FEC524', "#1D42BA", "#FFC72C",
                "#CE1141", "#FDBB30", "#1D428A", "#FDB927", "#5D76A9",
                "#98002E", "#00471B", "#78BE20", "#85714D", "#F58426",
                "#007AC1", "#0077C0", "#006BB6", "#E56020", "#E03A3E",
                "#5A2D81", "#C4CED4", "#CE1141", "#002B5C", "#E31837")

ggplot(data = contractdata, mapping = aes(fct_reorder(team, contractjudge), 
                                          contractjudge,
                                          fill = team)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_y_continuous(expand = c(0,0),
                     breaks = seq(-6, 5, 1)) +
  scale_fill_manual(values = teamcolors) +
  theme_minimal() +
  geom_text(aes(label = team), y = ifelse(contractjudge > 0, contractjudge - .5, contractjudge + .5),
            size = 3, angle = 90, color = ifelse(contractjudge == 0, 'black', 'white')) +
  labs(y = "Underpaid Players - Overpaid Players",
       title = "Contract Judge by NBA Team",
       x = NULL) +
  theme(axis.text.x = element_blank())
```


## Graph 3
```{r echo = FALSE, message = FALSE}
winandcontract <- merge(contractdata, winpct, by = "team")

ggplot(winandcontract, mapping = aes(contractjudge, winpct)) +
  geom_point(aes(color = team), size = 3, show.legend = FALSE) +
  geom_smooth(method = lm, se = F) +
  scale_color_manual(values = teamcolors) +
  annotate(geom = "text", x = 0.5, y = 0.826, size = 3.5, label = "GSW", color = "#FFC72C") +
  annotate(geom = "text", x = 3.5, y = 0.826, size = 3.5, label = "PHO", color = "#E56020") +
  annotate(geom = "text", x = -5.6, y = 0.417, size = 3.5, label = "SAC", color = "#5A2D81") +
  annotate(geom = "text", x = 3.4, y = 0.625, size = 3.5, label = "MIL", color = "#00471B") +
  annotate(geom = "text", x = 4.5, y = 0.696, size = 3.5, label = "UTA", color = "#002B5C") +
  annotate(geom = "text", x = -2.5, y = 0.182, size = 3.5, label = "DET", color = "#1D42BA") +
  annotate(geom = "text", x = -2.4, y = 0.667, size = 3.5, label = "CHI", color = "#CE1141") +
  annotate(geom = "text", x = -0.55, y = 0.696, size = 3.5, label = "BRK", color = "#777D84") +
  annotate(geom = "text", x = 4.5, y = 0.269, size = 3.5, label = "NOP", color = "#85714D") +
  annotate(geom = "text", x = -3.5, y = 0.208, size = 3.5, label = "ORL", color = "#006BB6") +
  theme_minimal() +
  labs(x = 'Contract Judge',
       y = '2021-22 Win Percentage (through 25 games)',
       title = "Win Percentage vs. Contract Judge - NBA Teams")
  
```


## Graph 4
```{r echo = FALSE, message = FALSE}
ggplot(data = fullnbadata, mapping = aes(x = reorder(gooddeal, salary2021_22),
                                         y = salary2021_22)) +
  geom_bar(stat = "summary", aes(fill = gooddeal), show.legend = FALSE) +
  scale_y_continuous(labels = scales::dollar,
                     expand = c(0,0)) +
  scale_fill_manual(values = c("#0e09fa", "#0076ff", "#00a9ff", "#00d3ff", "#07f6eb")) +
  theme_classic() +
  labs(x = 'Contract Type',
       y = 'Average Salary',
       title = 'Average Salary by Contract Type')
```

## Graph 5
```{r echo = FALSE, message = FALSE}
ggplot(data = fullnbadata, mapping = aes(x = reorder(gooddeal, war_total),
                                         y = war_total)) +
  geom_bar(stat = "summary", aes(fill = gooddeal), show.legend = FALSE) +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  scale_fill_manual(values = c("#2afa09", "#22e008", "#1bc707", "#14af06", "#0e9705")) +
  labs(x = 'Contract Type',
       y = 'Average WAR',
       title = 'Average WAR by Contract Type')
```


## Graph 6
```{r echo = FALSE}
orderednba = arrange(fullnbadata, difference)
woats = head(orderednba, n = 5)
goats = tail(orderednba, n = 5)

top_bottom_5 = rbind(woats, goats)

ggplot(data = top_bottom_5, mapping = aes(fct_reorder(player, difference),
                                          difference,
                                          fill = team)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = gsub(" ", "\n", player)), size = 2.9,
            y = ifelse(top_bottom_5$difference > 0,
                       top_bottom_5$difference - .5,
                       top_bottom_5$difference + 1)) +
  geom_text(aes(label = paste0("(", team, ")")), size = 2.9,
            y = ifelse(top_bottom_5$difference > 0,
                       top_bottom_5$difference - 1,
                       top_bottom_5$difference + .5)) +
  scale_fill_manual(values = c('green', 'red', 'green', 'green', 'red',
                               'red', 'red', 'red', 'green', 'green')) +
  theme_minimal() +
  labs(x = 'NBA Player',
       y = 'Scaled Difference Between Efficiency and Salary',
       title= 'Top 5 Worst and Best Contracts in the NBA') +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```


